---
title: "Code"
author: "Mahima Mago"
date: "2023-03-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploratory Aanlysis

```{r reading dataset, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
animal_shelter<-read.csv("dataset18.csv")
head(animal_shelter)
glimpse(animal_shelter)
```

Draw the probability density function of the number of days animal spends in the shelter.

```{r pdf of number of days animal spends in the shelter, include=TRUE}
ggplot(animal_shelter, aes(x = time_at_shelter)) + 
  geom_density(color = "blue")
```

Beacuse the response variable is the number of days animals spend in the shelter, which is integer data. In that case, we choose GLM model for Poisson distribution to fit the data.

```{r transform data format, include=FALSE}
animal_shelter$animal_type <- as.factor(animal_shelter$animal_type)
levels(animal_shelter$chip_status) <- c("BIRD", "DOG","CAT","WILDLIFE","LIVESTOCK")
animal_shelter$intake_type <- as.factor(animal_shelter$intake_type)
animal_shelter$outcome_type <- as.factor(animal_shelter$outcome_type)
animal_shelter$chip_status <- as.factor(animal_shelter$chip_status)
```


## Relationship between explanatory variables and response variable

Now, let's look at some plots of explanatory variable and response variable to get an initial impression of the data. In this dataset, animal_type, month, year, intake_type, outcome_type and chip_status are factors that have potential relationship with time_at_shelter, response variable in this dataset. Time_at_shelter indicates the number of days animals spend in the shelter.

Firstly, draw barchart and boxplot of animal_type and time_in _shelter to check if their have relationship.

```{r barchart and boxplot of animal_type and time_in _shelter, include=TRUE}
ggplot(data = animal_shelter, mapping = aes(x = animal_type,fill=animal_type))+
  geom_bar(stat="count", width= 0.5) +
  labs(x = "Animal types", y = "The number of days",
       title = "The number of days spent in the shelter of different animals") 
ggplot(data = animal_shelter, aes(x = animal_type, y = time_at_shelter, fill = animal_type)) +
  geom_boxplot() +
  labs(x = "The number of days spent in the shelter of different animals", y = "The number of days")+ 
  theme(legend.position = "none")
```

Here we can see that the dogs tend to spend the most time in shelter than that of other animals. And the birds and livestocks spend the least time in shelter. Also, there are small number of outliers in dog, cat and wildlife. Overall, the number of days of different animals spending in shelter are similar and most of them are stay below 20 days.

Let's look at the relationship between month and the number of days.

```{r factor month,include=FALSE}
animal_shelter$month
factor(animal_shelter$month)
```

```{r Barchart of month and time_in _shelter, include=TRUE}
ggplot(data = animal_shelter, mapping = aes(x = factor(month),fill=month))+
  geom_bar(stat="count", width= 0.5) +
  labs(x = "Month", y = "The number of days",
       title = "The number of days spent in the shelter of different month") +
  scale_x_discrete(labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                              "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))
```

From the barchart we can see that the shelter accepts the most animals in July and the least animals in February. This suggests that months may be related to the number of days animals spend in shelter.

What is the relationship between year and the number of days?

```{r Boxplot of year and time_in _shelter, include=TRUE}
animal_shelter$year <- as.factor(animal_shelter$year)
ggplot(data = animal_shelter, aes(x = year, y = time_at_shelter, fill = year)) +
  geom_boxplot() +
  labs(x = "The number of days spent in the shelter on different years", y = "The number of days")+ 
  theme(legend.position = "none")
```

We can see that the average number of days animals in shelter in 2016 is slightly larger than in 2017. But overall they are quite similar. There also have some outliers.

Next is the boxplot of intake_type and time_in _shelter.

```{r Boxplot of intake_type and time_in _shelter, include=TRUE}
ggplot(data = animal_shelter, aes(x = intake_type, y = time_at_shelter, fill = intake_type)) +
  geom_boxplot() +
  labs(x = "The number of days spent in the shelter of different intake types", y = "The number of days")+ 
  theme(legend.position = "none")
```
We see in the boxplot that the confiscated animals stayed in shelters the longest and the average number of days spent in shelter for animals abandoned by their owners was slightly lower than for lost animals. 

```{r Boxplot of chip_status and time_in _shelter, include=TRUE}
ggplot(data = animal_shelter, aes(x = chip_status, y = time_at_shelter, fill = chip_status)) +
  geom_boxplot() +
  labs(x = "The number of days spent in the shelter of different chip status", y = "The number of days")+ 
  theme(legend.position = "none")
```

The median number of days that animals with chips spent in shelter was almost the same as animals that without scanchips, and higher than animals that are unable to scan.

Now, let's fit a generalized linear model to see which factor is a significant predictor of the odds of the number of days animals spend in shelter.
